Waggle dance distributions quantify collective foraging in honey bee colonies

### 2022-07-26 analysis/manuscript/paper.Rmd
updated MS to new verion for current biology

The extraordinary waggle dance communication system of honeybees (Apis spp.) is a key example of how social behaviour can take on a form extending beyond that of the individual units [@Sumpter2006], such that groups function as collectives that make decisions as a single entity. Within colonies, a series of simple rules determine when and how much bees perform the waggle dance (Fig. 1) mean that choices between feeding sites occur at the level of the group rather than the individual [@Seeley1994]. For example, because the number of dance circuits performed by a bee on returning from a food source reflects the net energetic benefits of the trip, more of the colony's workforce will be recruited to the richer of two equidistant sources [@Seeley1991], or the closer of two equally rich sources [@Hasenjager2022]. This is without requiring any individuals to visit multiple resources to make a direct comparison. The resulting colony-level decisions are a prime example of how self-organized collective behaviour could increase the foraging efficiency of groups.

A wealth of research now shows that foraging bees frequently do not use dance information to find forage sites [@Gruter2009]. Accordingly, in many situations, colonies that are prevented from communicating via dances achieve equal or even greater foraging success than their wild-type counterparts [@Sherman2002;@Dornhaus2004;@Gruter2013;@Price2019]. This plasticity in the use of collective behaviour is intriguing because it may provide clues as to the selection pressures that were critical in the evolution of the dance communication system.

Identifying the circumstances in which bees use the dance is likely to be a key step towards understanding when it is likely to afford fitness benefits at the colony level. Despite the amount of research interest that has focused on the proximate mechanisms of collective decision-making via dance communication, the ultimate evolutionary drivers of its unique evolution within Apis -unparalleled in even a single other social insect- remain obscure [@Price2015]. While it is likely that the spatiotemporal distribution of forage was key, some studies find the benefits of dance communication to be realized only in challenging environments where resources are clumped or ephemeral [@Sherman2002;@Dornhaus2004;@Dornhaus2006], while others find no effects of landscape heterogeneity [@Nurnberger2017]. Others have found that dancing comes at a cost in challenging environments [@Price2019] and is most beneficial when species richness is high [@Donaldson-Matasci2012]. However, identifying how variation in landscape structure is associated with waggle dance use is practically difficult.

Attempts to identify the circumstances in which dance communication is important have been challenging because they require constant monitoring of real-world colony weight [@Sherman2002;@Dornhaus2004;Gruter20013;@Price2019;@Dornhaus2006;@Nurnberger2017;@Kirchner1998] -a noisy proxy of foraging success that is influenced by many other abiotic and biotic factors, including colony health- over ecologically credible time periods [@Schurch2014]. To do this at a scale that allows inter-colony variation in foraging environments at sufficient replication is a major logistical hurdle (but see [@Nurnberger2017]), particularly given that multiple landscape variables may interact to determine the utility of dance communication.

@@

Here, we present a method to determine the extent of waggle dance use in colony foraging within specific landscapes by identifying the proportion of foraging trips that are made by bees recruited through waggle dances. Our method examines the distribution of waggle run durations reported on the dance-floor and infers from its shape the relative contribution of dances for resources identified by individual search, “scouting” versus dances for resources found by following dances, “recruitment” [@Seeley1995]. Previously, the relative amount of scouting and recruitment trips have only been quantifiable by documenting foragers' individual search history, which requires tracking individual bees and is time -and labour- intensive [@Seeley1995]. Our methods provide an efficient means of quantifying collective behaviour with no requirement to manipulate hive orientations, assay foraging efficiency or document individuals' previous experience. As such, they can identify when colonies adopt a collective foraging strategy- a critical step towards understanding when such a strategy outperforms individual-level behaviour.

@@

The patterns of waggle dance-encoded foraging distances might differ under scenarios where colonies rely on individual search versus recruitment. To show this, we simulated honeybees foraging in a landscape where resource patches were randomly placed in the environment. Foragers could locate these under two different strategies: either acting as a scout and locating resources themselves, or following a recruit strategy and locating resources by following a random dance from the dance floor [@Seeley1992] (Fig. 2a,b, details in Methods). As it is known in the simulation which individuals in the hive forage under what strategy, we can compare the distributions of foraging distances reported on the dance floor by each type of forager. Fig 2c shows that the shapes of the resource distance distributions for bees engaging in the two types of foraging trips are different. The distance distribution for the scout trips is akin close to that of an exponential distribution (Fig. 2c), which is the nearest neighbour distance distribution for foragers operating in a one-dimensional environment (see methods). The distribution of the distances reported for recruit trips (Fig. 2c) is a Rayleigh distribution which is the nearest-neighbour distribution in a two-dimensional environment [@Pyke1978] (see Methods).

@@

On the dance-floor, the number of waggle runs performed for a resource depends on its profitability [@Seeley1994]. Honeybees achieve this by measuring the energetic efficiency of a foraging trip through the ratio of energetic gain to energetic cost [@Seeley1994;@Seeley1994A]. By combining this profitability bias with the distributions identified in our simulations, we can accurately describe the distribution of waggle runs reported on the dance-floors of real honeybee colonies as a superposition of scout and recruit distributions (Fig 3, see Methods for details)

@@

```{r, fig.cap="Figure 3. **The rationale of the foraging model**. The distances of resources encountered by scouts are distributed exponentially (a). These dances are advertised on the dance floor (b). Dances for resources that are closer or higher in quality are repeated more often (c). As a consequence, dances for more profitable resource are over-represented and sampling foragers are biased to the more profitable resources (d). After successfully visiting advertised resources, recruits also dance for them leading to further amplification of this bias towards the most profitable resource in the vicinity of the hive (e).  The distances of recruiting trips are than distributed through a Rayleigh distribution (f). The distances reported on the dance floor are a mixture of the scout and recruiting trips and can be calculated from the distance distributions of the scouting and recruiting trips, taking the reporting bias into account (see Methods for detail)."}
knitr::include_graphics("../results/figures/manuscript_figures/fig3.png")
```

To evaluate the use of waggle dance information and individual search in honeybee colonies foraging in 'natural' landscapes, we analysed a pre-existing dataset of 2827 waggle dance observations from 20 observation hives, recorded between April-September 2017, (previously described in [@Samuelson2021]). Hives that contributed to this dataset had been situated at different locations in South East England (see Methods, figure 4A) and visited every two weeks for a period of 24 weeks. On each visit, two hours of continuous waggle dance data was recorded by training a camcorder onto the dance floor. The footage of the dances was decoded manually [@Couvillon2012;@Schurch2019;@Samuelson2021] to extract waggle run durations. Using this data, for each site we fit both the collective and individual models and used model selection to determine which provided the better explanation of the data, and (if the collective model provided a better fit) to quantify the relative use of social information through estimating the parameter $p$. In each case, We calculated the goodness-of-fit using a Kolmorgorov-Smirnov (KS) test to ascertain if the model provided a plausible explanation of the data [@Goldstein2004;@Clauset2009].

For 16 out of 20 study hives, the collective model provided a better description of the data than the individual model (Fig. 4a). In the other 4 sites, despite the collective model having the higher maximum likelihood, the individual model had a higher AIC value and so is more parsimonious (Supplementary Table 1). In all but one site, the collective model had a good fit (using a Kolmogorov-Smirnov statistic of $p > 0.05$, see methods) to the empirical waggle run durations (Fig. 4b), whereas the individual model was significantly different to the observed data in 8 sites (Kolmogorov-Smirnov statistic $p < 0.05$, Fig. 4b). The sites shown in Figs 4c-d are representative examples showing the model fits where the individual (Fig. 4c) and collective (Fig. 4d) models fit best. Note the closeness of the fit to the data, illustrating the overall quality of the model description.

@@

Further quantification of the use of waggle-dance recruitment within all colonies, as a proportion of all foraging trips, can be achieved by extracting the estimated proportion of scout trips, $p$, for each site. Since our sites varied in land-use characteristics and potentially thus forage distributions (although in this case, not by design for this study), we investigated whether these estimates might correlate with land-use. We first classified the different land-use types of the area surrounding each site [@Samuelson2018] to obtain a standardised land-use profile for the urban and agri-rural environments separately as many land-use types present in urban areas do not occur in agri-rural environments and vice versa (see methods). We then performed a Partial Least Squares (PLS) analysis [@Carrascal2009] (see methods) to determine the principal components that represent combinations of land-use types which explained the most variation in the proportion of scout dances within agri-rural and urban environments. As for one of our sites neither model provided a plausible description, this site was removed from the PLS analysis. Due to our small sample size (10 urban and 9 agri-rural sites), we used jackknife resampling to evaluate the robustness of our results to influential points (see Methods, Supplementary Material).

@@

# Discussion

Wwe have presented a model to quantify the use of waggle-dance communication in collective decision-making by honeybee colonies. Whilst recruitment is well known to occur through other mechanisms besides the waggle dance, such as through olfactory cues [@Arenas2007], the exceptionally close fit of our model to waggle run durations underlines the importance of the waggle dance in honey bee foraging. The fitting of our model to waggle dance observations provides a time and labour efficient methodology to quantify collective behaviour. Our model provides a tool to map the environment along the major axis of honeybee information use and visualise how land-use influences the use of the waggle dance and thus identifies the land-use combinations which most influence variation in the proportion of scouts.

Whilst our results show a decrease in reliance on waggle dance recruitment as resources become harder to find and foraging trips become longer [@Samuelson2021], they do not yet provide conclusive evidence for such a pattern. Although the agri-rural results agree with this assessment, the urban results seem to go against this trend. As we used an existing dataset, we did not systematically choose sites based on forage availability, and instead inferred forage availability based on land-use at each site. The dance recordings in our dataset were collected over an extended period of five months, over which time forage availability in the landscape likely changed considerably and non-uniformly across sites. We thus cannot rule out that longer term effects not captured in the data used in this study, such as resource stability [@Schurch2014], may also have contributed to our estimates of waggle dance use. Our findings illustrate how our model can be used to estimate reliance on collective behaviour within different landscapes, for exploration within datasets where land-use is systematically chosen by design.

Our results within the agri-rural sites are in agreement with other studies which evaluated foraging performance with and without the dance [@Dornhaus2006;@Price2019]. These results suggest that, even though collective foraging is not always beneficial, in environments where high quality resources are present, but relatively scarce, collective foraging is being carried out by colonies, suggesting that under these conditions exchanging social information through the waggle dance confers benefits. In both agri-rural and urban landscapes we have shown that the proportion of scouting trips change with land-use (illustrated in the change in land-use shown in the maps from left to right in Fig 5). We arrived at this conclusion through the analysis of waggle dance data, using a mathematical model to fit the waggle dance durations against. This method provides a time and labour efficient methodology to quantify collective behaviour. We analysed an existing data set and thus provide a proof of concept for how this new toolkit can be used to evaluate factors influencing waggle dance recruitment. The methods described here can also provide a framework to start exploring how collective decision making can help to buffer the effects of landscape change, and in this case how flexibility in the use of the waggle dance may help to exploit resources when experiencing novel landscape mosaics such as in the highly urbanised conurbations.

The analysis of waggle dance data can make an important contribution to our understanding of social information use and provides a methodology to further evaluate how honey bees use their unique dance language. Recently, technological advances have emerged which enable colony metrics to be collected faster, more accurately and over greater time spans than could be gathered by hand, allowing individuals to be tracked within colonies and theories of individual behaviour to be evaluated in more depth than could have been done previously [@Wild2021]. By piecing together the behavioural response of individuals and combining these with landscape analyses, we have found an accurate mathematical description of colony foraging which extends our ability to quantify collective behaviour across environments. With the advances in the decoding of the waggle dance through automated methods [@Wario2017], we face the prospect of waggle dance data becoming “big data”. Our methodology thus provides a means of analysing such large data sets to inform the debate about the importance of collective decision making, as well as providing useful colony metrics of foraging activity.

@@

Methods for how the data used for this study was collected can be found in full in Materials and Methods sections 2.1, 2.2 and 2.3 of [@Samuelson2021].

### Land-use preference analysis

Methods for how land-use types were classified can be found in Materials and Methods sections 2.6 of [@Samuelson2021].

### Waggle dance decoding

Methods for waggle dance decoding are fully described in Materials and Methods section 2.4 in [@Samuelson2021].

@@

A circular environment is first created with radius $r = 2.5$. The number of resources in the environment is generated as a random Poisson variable with rate equal to 5000 multiplied by the area of the environment. These are placed on polar coordinates with a uniformly selected angle, $\theta$, between 0 and 2$\pi$ and a radial value, $\rho$, between 0 and $r$, determined from the square root of the uniform position values multiplied by $r$. These polar coordinates are converted to Cartesian coordinates. Each location is then assigned to an instance of a resource object along with a random quality of between 0 and 10. This quality is combined with the distance of the resource to the centrally located hive to form a measure of how profitable the resource is (see model, equation?).

One-hundred honeybee objects are created, 20 of which 20 are on scouting trips and the rest recruited to follow scout dances. Scouts leave the environment following a random path through the environment generated by sampling a uniform random step length and angle. The number of paths the scout draws when searching is also determined as a uniform random number. Each straight line in the random path is converted to a rectangle with length equal to the path section length and a constant width of ~0.01 to represent an area the scout searches along that path. Of all the resources contained in the boxes drawn from the scout's path, the one closest to the colony is selected as the resource patch that the scout will report and will communicate its location if the quality of the resource resource exceeds a minimum threshold. Communication is simulated by pooling together all the resource patches found. If no resources are contained in the scout's path, they will not add any resources to the scout pool and draw a new path in the next foraging iteration.

@@

To describe the distribution of dance durations on the dance floor we formulated a generic model the duration of waggle dances. In the model resource patches are assumed to be randomly placed in the environment. Foragers scout for these patches. The rationale of the model is illustrated in Fig. 3: upon visiting a resource patch, foragers translate the profitability of a resource into the number of repeats of the dance. The number of repeats of the dance is a function of quality an distance. Recruits sample random dances and report the location of successful visits to resource patches on the dance floor. Through the feedback and over-representation of profitable resources on the dance floor recruits will converge to visiting the most profitable resource in vicinity of the hive. The distribution of dance durations is the superposition of scouting and recruiting trips.

@@

Prior to conducing the PLS we removed any sites in which the models fit was significantly different to the waggle dance durations for that site. This resulted in one agri-rural site (ROT) and no urban sites being removed from the analysis.

@@

All data and code is available on zenodo at....

>>>>

In group living animals, many decisions are taken collectively by integrating information from individuals to produce a social behaviour which extends beyond that of the individual units [@Sumpter2006]. Such systems are self-organised, relying on the constituents using socially provided information and following behavioural 'rules' to produce a collective outcome [@Couzin2003;@Bonabeau1997]. However, in addition to information provided by others, individuals also have knowledge of their own (private information) and so can choose to work collectively or individually [@Dall2005;@Reiucau2011;@Gruter2014]. The benefits of working collectively versus independently vary across environmental contexts and individuals often adjust their own use of information accordingly [@Dechaume-Moncharmont2005;@Beekman2008]. The impact these individual variations in information use have at a group level, however, is less well understood and difficult to empirically evaluate [@Rieucau2011;@Giardina2008;@Delgado2018].

Understanding the collective ramifications of individual variation in information use is not only important for deciphering how social animals collectively regulate information use [@Gruter2014], relating this to the environmental context of the behaviour is key to illuminating the ultimate evolutionary drivers of collective processes [@Price2015]. However, establishing the group level effects of individual variation in information use is not trivial and requires measuring information use across animal groups, an often laborious process which requires monitoring individuals' behaviours and interactions. This becomes logistically impossible in very large animal groups, such as in social insects, which display some of the most intricate group decision-making mechanisms and consist of often thousands of individuals, each using a mix of individual and collective information to make decisions [@Bonabeau1997].

Social insects are defined by their highly coordinated collective responses and so provide ideal systems to investigate the impact of individual variation in information use on emergent collective processes. Nevertheless, the sheer quantity of individuals in many colonies is a barrier to further understanding these dynamics. Honey bees (Apis spp.) provide an opportunity to sidestep this problem of scale as they provide a detailed record of colony foraging activity in the form of their unique waggle dance -unparalleled in even a single other social insect [@Seeley1995]. Using the dance, individuals share information about profitable resource locations. A series of simple rules which determine when and how much bees perform waggle dances (Fig. 1) mean that choices between feeding sites occur at the level of the group rather than the individual [@Seeley1991;@Seeley2004;@Hasenjager2022]. Despite this, a wealth of research shows that individual foragers frequently do not use dance information to find forage sites and instead rely on their own knowledge [@Gruter2009]. Moreover, studies investigating the value of the dance have shown that when prevented from communicating via dances, honey bee colonies can achieve equal or even greater foraging success than their wild-type counterparts [@Sherman2002;@Dornhaus2004;@Gruter2013;@Price2019].

@@

The waggle dance provides a model system to study the group effects of collective and individual information use. Although it is likely that individual variation in dance use translates into variations in collective decision making at the colony level [@Donaldson-Matasci2012;@Price2019], this dynamic has not been empirically established. Individual based models, which explore the collective effects of individuals following different information streams provide an insight into how individual behaviours scale to the level of the group [@Dornhaus2006;Beekman2008], however, these are difficult to empirically relate to real observations and so are limited to showing what a colony under certain conditions might do, rather than quantifying what they are doing. Empirically determining how dance information is used at a colony level requires new methods which can identify the behaviours of individual foragers and integrate them with the mechanisms by which the dance functions to make collective decisions. Previously, this has only been quantifiable by documenting foragers' individual search history, which requires tracking individual bees and is time -and labour- intensive [@Seeley1995].

Here, we investigate how variation in individual information use influences the collective foraging behaviour of honey bees. Using a model based on well-established foraging behaviours [@Seeley1986;@Seeley1991;@Seeley1992;@Seeley1994;@Seeley1994A;@Seeley1995], we examine the distribution of waggle run durations and infer from its shape the relative contribution of dances for resources identified by individual search - “scouting”- versus dances for resources found by following dances - “recruitment”- [@Seeley1995]. This allows us to quantify the extent to which colonies use dance communication and thus forage collectively. To investigate how information use varies across environment, we fit our model to waggle dance observations from 20 sites in an agri-rural and urban environment (10 in each) and evaluated the landscape characteristics which best explained the variance in dance use.

@@

The patterns of waggle dance-encoded foraging distances might differ under scenarios where colonies rely on individual search versus recruitment. To test this, we simulated honeybees foraging in a landscape where resource patches were randomly placed in the environment. Foragers could locate these under two different strategies: either acting as a scout and locating resources themselves, or following a recruit strategy and locating resources by following a random dance from the dance floor [@Seeley1992] (Fig. 2a,b, details in Materials and Methods). As it is known in the simulation which individuals in the hive forage under what strategy, we can compare the distributions of foraging distances reported on the dance floor by each type of forager. Fig 2c shows that the shapes of the resource distance distributions for bees engaging in the two types of foraging trips are different. The distance distribution for the scout trips is akin close to that of an exponential distribution (Fig. 2c), which is the nearest neighbour distance distribution for foragers operating in a one-dimensional environment (see Materials and Methods). The distribution of the distances reported for recruit trips (Fig. 2c) is a Rayleigh distribution which is the nearest-neighbour distribution in a two-dimensional environment [@Pyke1978] (see Materials and Methods).

@@

On the dance-floor, the number of waggle runs performed for a resource depends on its profitability [@Seeley1994]. Honeybees achieve this by measuring the energetic efficiency of a foraging trip through the ratio of energetic gain to energetic cost [@Seeley1994;@Seeley1994A]. By combining this profitability bias with the distributions identified in our simulations, we can accurately describe the distribution of waggle runs reported on the dance-floors of real honeybee colonies as a superposition of scout and recruit distributions (Fig 3, see Materials and Methods for details)

@@

```{r, fig.cap="Figure 3. **The rationale of the foraging model**. The distances of resources encountered by scouts are distributed exponentially (a). These dances are advertised on the dance floor (b). Dances for resources that are closer or higher in quality are repeated more often (c). As a consequence, dances for more profitable resource are over-represented and sampling foragers are biased to the more profitable resources (d). After successfully visiting advertised resources, recruits also dance for them leading to further amplification of this bias towards the most profitable resource in the vicinity of the hive (e).  The distances of recruiting trips are than distributed through a Rayleigh distribution (f). The distances reported on the dance floor are a mixture of the scout and recruiting trips and can be calculated from the distance distributions of the scouting and recruiting trips, taking the reporting bias into account (see Materials and Methods for detail)."}
knitr::include_graphics("../results/figures/manuscript_figures/fig3.png")
```

To evaluate the use of waggle dance information and individual search in honeybee colonies foraging in 'natural' landscapes, we analysed a pre-existing dataset of 2827 waggle dance observations from 20 observation hives, recorded between April-September 2017, (previously described in [@Samuelson2021]). Hives that contributed to this dataset had been situated at different locations in South East England (see Materials and Methods, figure 4A) and visited every two weeks for a period of 24 weeks. On each visit, two hours of continuous waggle dance data was recorded by training a camcorder onto the dance floor. The footage of the dances was decoded manually [@Couvillon2012;@Schurch2019;@Samuelson2021] to extract waggle run durations. Using this data, for each site we fit both the collective and individual models and used model selection to determine which provided the better explanation of the data, and (if the collective model provided a better fit) to quantify the relative use of social information through estimating the parameter $p$. In each case, We calculated the goodness-of-fit using a Kolmorgorov-Smirnov (KS) test to ascertain if the model provided a plausible explanation of the data [@Goldstein2004;@Clauset2009].

For 16 out of 20 study hives, the collective model provided a better description of the data than the individual model (Fig. 4a). In the other 4 sites, despite the collective model having the higher maximum likelihood, the individual model had a higher AIC value and so is more parsimonious (Supplementary Table 1). In all but one site, the collective model had a good fit (using a Kolmogorov-Smirnov statistic of $p > 0.05$, see Materials and Methods) to the empirical waggle run durations (Fig. 4b), whereas the individual model was significantly different to the observed data in 8 sites (Kolmogorov-Smirnov statistic $p < 0.05$, Fig. 4b). The sites shown in Figs 4c-d are representative examples showing the model fits where the individual (Fig. 4c) and collective (Fig. 4d) models fit best. Note the closeness of the fit to the data, illustrating the overall quality of the model description.

@@

Further quantification of the use of waggle-dance recruitment within all colonies, as a proportion of all foraging trips, can be achieved by extracting the estimated proportion of scout trips, $p$, for each site. Since our sites varied in land-use characteristics and potentially thus forage distributions (although in this case, not by design for this study), we investigated whether these estimates might correlate with land-use. We first classified the different land-use types of the area surrounding each site [@Samuelson2018] to obtain a standardised land-use profile for the urban and agri-rural environments separately as many land-use types present in urban areas do not occur in agri-rural environments and vice versa (see Materials and Methods). We then performed a Partial Least Squares (PLS) analysis [@Carrascal2009] (see Materials and Methods) to determine the principal components that represent combinations of land-use types which explained the most variation in the proportion of scout dances within agri-rural and urban environments. As for one of our sites neither model provided a plausible description, this site was removed from the PLS analysis. Due to our small sample size (10 urban and 9 agri-rural sites), we used jackknife resampling to evaluate the robustness of our results to influential points (see Materials and Methods, Supplementary Material).

@@

# Discussion New

Understanding how individual variation in information use scales to produce flexible collective processes at the group level is a key challenge in behavioural ecology. Here, we have presented a method to quantify information use at a collective level in a model social insect. Using this method, we have shown that collective decision-making is used variably. Whilst recruitment is well known to occur through other mechanisms besides the waggle dance, such as through olfactory cues [@Arenas2007], the exceptionally close fit of our model to waggle run durations underlines the importance of the waggle dance in honey bee foraging. Further examining our estimates of dance use show that a colonies use of the waggle dance varies across environments and does not appear to be randomly determined.

Within both the agri-rural and urban environments, dance use had a significant interaction with land-use type, ranging from total dance use to complete reliance on individual information. The notion that individual variation in information use influences the collective effect has only previously been inferred theoretically. Our findings demonstrate this empirically. Individual-based models [@Dornhaus2006;Beekman2008;@Price2019] have outlined how variations at an individual level could drive group level dynamics, however, without verifying this upon empirical data the insights remain theoretical. Here, we have shown how an individual based model can be used to better understand a system in a way which allows us to build more informative models which can be empirically verified. Not only do our findings add support for our understanding of how the dance functions to make collective decisions [@Seeley1995], but they also provide a new dimension to our understanding of how variation at an individual level influences the collective level.

Whilst the benefits of dance communication are known to vary across environments [@Donaldson-Matasci2012;@Gruter2011;@Schurch2014;@Price2019], our results are the first to show colonies vary it's use in response, affording the colony a flexible dance use strategy which they can tailor to different conditions to maximise foraging success. Identifying the circumstances in which dance communication is important have been hampered by the requirement to monitor real-world colony weight [@Sherman2002;@Dornhaus2004;Gruter2013;@Price2019;@Dornhaus2006;@Nurnberger2017;@Kirchner1998] -a noisy proxy of foraging success that is influenced by many other abiotic and biotic factors, including colony health- over ecologically credible time periods [@Schurch2014]. Doing this at a scale that allows inter-colony variation in foraging environments at sufficient replication is a major logistical hurdle (but see [@Nurnberger2017]), particularly given that multiple landscape variables may interact to determine the utility of dance communication. Our method circumvents this issue by providing a time and labour efficient methodology to quantify this key collective behaviour and map the environment along the major axis of honeybee information use.

The methods described here provide a framework to start exploring how plasticity in collective decision making can help to buffer the effects of landscape change, and in this case, how flexibility in the use of the waggle dance may help to exploit resources when experiencing novel landscape mosaics, such as in highly urbanized conurbations. As we used an existing dataset, we did not systematically choose sites based on forage availability, and instead inferred forage availability based on land-use at each site. The dance recordings in our dataset were collected over an extended period of five months, over which time forage availability in the landscape likely changed considerably and non-uniformly across sites. We thus cannot rule out that longer term effects not captured in the data used in this study, such as resource stability [@Schurch2014], may also have contributed to our estimates of waggle dance use. The number of colonies used in our analysis, although comparable to studies in which the benefits of collective foraging have been evaluated [@Sherman2002;@Dornhaus2004;@Price2019;@Nurnberger2017;Kirchner1998], limits the statistical power of our analysis. Despite this, we gain useful insights into the drivers of this relationship by exploring how land-use type influences waggle dance use.

Our agri-rural results suggest a potential decrease in reliance on waggle dance recruitment as resources become harder to find and foraging trips become longer [@Samuelson2021]. Our results within the agri-rural sites broadly agree with the findings from other theoretical [@Dornhaus2006] (11) and empirical [@Price2019] studies which examined foraging performance with and without the dance. These results suggest that, even though collective foraging is not always beneficial, in environments where high quality resources are present but relatively scarce, collective foraging is being carried out by colonies and under these conditions exchanging social information through the waggle dance confers benefits. In both agri-rural and urban landscapes we have shown that the waggle dance use changes with land-use (illustrated in the change in land-use shown in the maps from left to right in Fig 5A and 5C). These findings illustrate how our model can be used to estimate reliance on collective behaviour within different landscapes, for exploration within datasets where land-use is systematically chosen by design.

The analysis of waggle dance data can make an important contribution to our understanding of information use. Recently, technological advances have emerged which enable colony metrics to be collected faster, more accurately and over greater time spans than could be gathered by hand, allowing individuals to be tracked within colonies and theories of individual behaviour to be evaluated in more depth than could have been done previously [@Wild2021]. By piecing together the behavioural response of individuals and combining these with landscape analyses, we have found an accurate mathematical description of colony foraging which extends our ability to quantify collective behaviour across environments. With the advances in the decoding of the waggle dance through automated methods [@Wario2017], we face the prospect of waggle dance data becoming “big data”. Our methodology thus provides a means of analysing such large data sets to inform the debate about the importance of collective decision making, as well as providing useful colony metrics of foraging activity.

@@

Details of data collection , waggle dance decoding and classification of land-use types can be found in full in the Materials and Methods section of [@Samuelson2021].

@@

A circular environment is first created with radius $r = 2.5$. The number of resources in the environment is generated as a random Poisson variable with rate equal to 5000 multiplied by the area of the environment. These are placed on polar coordinates with a uniformly selected angle, $\theta$, between 0 and 2$\pi$ and a radial value, $\rho$, between 0 and $r$, determined from the square root of the uniform position values multiplied by $r$. These polar coordinates are converted to Cartesian coordinates. Each location is then assigned to an instance of a resource object along with a random quality of between 0 and 10. This quality is combined with the distance of the resource to the centrally located hive to form a measure of how profitable the resource is (see model, equations).

One-hundred honeybee objects are created, 20 of which are on scouting trips and the rest recruited to follow scout dances. Scouts leave the environment following a random path through the environment generated by sampling a uniform random step length and angle. The number of paths the scout draws when searching is also determined as a uniform random number. Each straight line in the random path is converted to a rectangle with length equal to the path section length and a constant width of ~0.01 to represent an area the scout searches along that path. Of all the resources contained in the boxes drawn from the scout's path, the one closest to the colony is selected as the resource patch that the scout will report and will communicate its location if the quality of the resource resource exceeds a minimum threshold. Communication is simulated by pooling together all the resource patches found. If no resources are contained in the scout's path, they will not add any resources to the scout pool and draw a new path in the next foraging iteration.

@@

To describe the distribution of dance durations on the dance floor we formulated a generic model for the duration of waggle dances. In the model resource patches are assumed to be randomly placed in the environment. Foragers scout for these patches. The rationale of the model is illustrated in Fig. 3: upon visiting a resource patch, foragers translate the profitability of a resource into the number of repeats of the dance. The number of repeats of the dance is a function of quality an distance. Recruits sample random dances and report the location of successful visits to resource patches on the dance floor. Through the feedback and over-representation of profitable resources on the dance floor recruits will converge to visiting the most profitable resource in vicinity of the hive. The distribution of dance durations is the superposition of scouting and recruiting trips.

@@

Prior to conducing the PLS we removed any sites in which the models fit was significantly different to the waggle dance durations for that site. This resulted in one agri-rural site and no urban sites being removed from the analysis.

@@

All data and code is available on request and will be made available on zenodo prior to publication.
